Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-08-31 Report/Sections/appendix.tex
Updates from Overleaf

\begin{enumerate}
    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$:
    $$
    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(i^*)}\right)^2 \right)
    $$
    where $\sigma_A^{*(1)} = 0.05s$ and $\sigma_A^{*(2)} = 0.1s$. 
    %
    \item Next, for each dive $t$, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots ~ $ using Equation (\ref{eqn:yhat_0}) below:
    \begin{equation}	
        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
    	\label{eqn:yhat_0}
    \end{equation}
    where $\phi_A^{*(1)} = 0.99$ and $\phi_A^{*(2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
    %
    \item We then simulate Fourier modes 1 through 49 within each window:
    $$
    	\hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
    $$
    where $a^{(k)}_{t,t^*}$ is equals either 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
    \begin{align}
    \begin{split}
    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
    	%
    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
    	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
    	{\rm{Gamma}}(250,1), & k = 1 \\
    	{\rm{Gamma}}(50,1), & k = 2.
    	\end{array}\right. 
    \end{split}
    \end{align}
    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ Than subdive state 1 $(X^*_{t,t^*} = 1)$. Note that $\mathbb{E}\left(|\hat{Y}^{(k)}_{t,t^*}|^2\right)$ is proportional to $1/k^2$. This silences the higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$. 
    \item For Fourier modes 50 through 99, we use the following values to ensure that the inverse DFT is real-valued:
    $$
    \hat{Y}^{*(50)}_{t,t^*} = 0; \qquad
	\hat{Y}^{*(k)}_{t,t^*} = -\hat{Y}^{*(100-k)}_{t,t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
    $$
    %
    \item Finally, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
    %
    $$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
\end{enumerate}
%
%$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window. $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated. See Figure \ref{fig:sim_data} for details.

We now show that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}. First, note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$ because the real part of the Fourier coefficient corresponding to a frequency of 0 is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction. Then, combine this fact with Equation (\ref{eqn:yhat_0}) to find the distribution of $\Zone_{t,t^*}$: 
%
$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right).$$
%
This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$. We also have that $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.

To calculate the wiggliness $\Ztwo$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
%
$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\hat{Y}^{(k)}_{t,t^*}\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
Combining this with Equation block (5), we have:
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right)$$
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right).$$
%
It follows that $\mu_W^{*(1)} = 10.10$, $\sigma_W^{*(1)} = 3.18$, $\mu_W^{*(2)} = 305.94$, and $\sigma_W^{*(2)} = 17.49$.

>>>>

\begin{enumerate}
    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$:
    $$
    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
    $$
    where $\sigma_A^{*(\cdot,1)} = 0.05s$ and $\sigma_A^{*(\cdot,2)} = 0.1s$. 
    %
    \item Next, for each dive $t$, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots ~ $ using Equation (\ref{eqn:yhat_0}) below:
    \begin{equation}	
        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
    	\label{eqn:yhat_0}
    \end{equation}
    where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
    %
    \item We then simulate Fourier modes 1 through 49 within each window:
    $$
    	\hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
    $$
    where $a^{(k)}_{t,t^*}$ is equals either 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
    \begin{align}
    \begin{split}
    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
    	%
    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
    	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
    	{\rm{Gamma}}(250,1), & k = 1 \\
    	{\rm{Gamma}}(50,1), & k = 2.
    	\end{array}\right. 
    \end{split}
    \end{align}
    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ Than subdive state 1 $(X^*_{t,t^*} = 1)$. Note that $\mathbb{E}\left(|\hat{Y}^{(k)}_{t,t^*}|^2\right)$ is proportional to $1/k^2$. This silences the higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$. 
    \item For Fourier modes 50 through 99, we use the following values to ensure that the inverse DFT is real-valued:
    $$
    \hat{Y}^{*(50)}_{t,t^*} = 0; \qquad
	\hat{Y}^{*(k)}_{t,t^*} = -\hat{Y}^{*(100-k)}_{t,t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
    $$
    %
    \item Finally, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
    %
    $$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
\end{enumerate}
%
%$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window. $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated. See Figure \ref{fig:sim_data} for details.

We now show that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}. First, note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$ because the real part of the Fourier coefficient corresponding to a frequency of 0 is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction. Then, combine this fact with Equation (\ref{eqn:yhat_0}) to find the distribution of $\Zone_{t,t^*}$: 
%
$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
%
This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,i^*)} = 0$. We also have that $\sigma_A^{*(\cdot,1)} = 0.05s$, $\phi_A^{*(\cdot,1)} = 0.99$, $\sigma_A^{*(\cdot,2)} = 0.1s$, and $\phi_A^{*(\cdot,2)} = 0.95$.

To calculate the wiggliness $\Ztwo_{t,t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
%
$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\hat{Y}^{(k)}_{t,t^*}\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
Combining this with Equation block (5), we have:
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right)$$
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right).$$
%
It follows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
